Local wind regime induced by giant linear dunes: comparison of ERA5 re-analysis with surface measurements

### 2022-02-28 Paper/main.tex
Entering Giles 'white' comments for abstract and introduction:
- Ignoring 'Arid areas'
- chossing something else for 'except for some weak winds blowing at
  night'

Keep inluding Giles comments and modifications:

- White comments until section3 (until page 9 included)
- changing ref to SI to 'Online Ressource ...' according to BLM template
- changing station names to 'North Sand Sea' and 'South Sand Sea' in
  text and figures

Changing title to Giles suggestion, and correction Mat affiliation

updating latexdiff before sending to Philippe

Phillipe corrections afetr Giles remark

Adding arro

\title{Local wind regime induced by giant linear dunes}

@@

\institute{
C. Gadal \at
Institut de M\'ecanique des Fluides de Toulouse, Universit\'e de Toulouse Paul Sabatier, CNRS, Toulouse INP-ENSEEIHT, Toulouse, France. \email{cyril.gadal@imft.fr}
\and
P. Delorme \at
School of Geography and Environmental Science, University of Southampton, Southampton, UK.
\and
C. Narteau \at
Institut de Physique du Globe de Paris, Universit\'e de Paris, CNRS, Paris, France.
\and
G.S.F. Wiggs \at
School of Geography and the Environment, University of Oxford, Oxford, UK.
\and
M. Baddock \at
Geography and Environment, Loughborough University of Technology, Loughborough, UK.
\and
J.M. Nield \at
School of Geography and Environmental Science, University of Southampton, Southampton, UK.
\and
P. Claudin \at
Physique et M\'ecanique des Milieux H\'et\'erog\`enes, CNRS, ESPCI Paris, PSL Research University, Universit\'e de Paris, Sorbonne Universit\'e, Paris, France.  \email{philippe.claudin@espci.fr}
}

@@

\begin{abstract}
Emergence and growth of sand dunes result from the interaction between the topography, which perturbs the wind flow, which itself controls the sediment transport. While these feedbacks are well studied at the scale of a single dune, the average effect of a periodic dune pattern on atmospheric flows remains poorly constrained due to a lack of data in major sand seas. Here, we compare field measurements of surface wind data to the predictions of the ERA5-Land climate reanalysis in four different places in Namibia, inside and north of the giant-dune field of the Namib sand sea. In the flat desert areas to the north, observations and predictions agree well despite the small amount of data in the assimilation process. This is also the case in the interdune areas of the sand sea, except for the weak winds blowing at night in a direction oblique to the orientation of the giant dunes. We quantify these similarities and differences and provide a physical understanding of the relevant aerodynamical regimes to relate them to the daily cycle of the turbulent atmospheric boundary layer over a dune pattern of given wavelength. We conclude by identifying the conditions under which the ERA5-Land reanalysis data can or can't be used to study dune morphodynamics. We also propose that, in multidirectional wind regimes, deflections of specific winds could explain the occurrence of secondary dune patterns with a different orientation than the primary structures between which they grow.
\keywords{Atmospheric boundary layer \and Sand dunes \and Fluid-structure interactions}
\end{abstract}

@@

The description of turbulent flows over a complex topography is a rich and active subject, that is relevant for a large variety of different environmental systems \citep{Finnigan2020}. For example, the flow over hills is of primary interest for wind power, meteorological and air pollution phenomena \citep{Taylor1987}. The properties of these flows are also key to the understanding of the formation of ocean surface wind-driven waves \citep{Sullivan2010}, dissolution bedforms \citep{Claudin2017}, or sedimentary ripples and dunes \citep{Charru2013, Courrech2015}. Importantly, the troposphere presents a vertical structure, with a lower convective boundary layer, of typical kilometer-scale thickness, capped by a stably stratified region \citep{Stull1988}. The largest topographic obstacles as mountains can therefore interact with this upper region, for example leading to internal wave generation or significant wind disturbances, such as downslope winds in lee sides \citep{Durran1990}.

Looking at the wind close to the surface, two topographic feedbacks on the flow, although related, can be commented separately. There is a first effect on the wind intensity: the flow accelerates on the upwind slope and slows down on the downwind one \citep{Baddock2011}, with a speed-up factor essentially proportional to the obstacle aspect ratio \pc{here a ref?}. Importantly, the velocity maximum is typically shifted upwind of the obstacle crest. This behaviour has been theoretically predicted by means of asymptotic analysis of a neutrally stratified boundary-layer flow over an obstacle of vanishing aspect ratio \citep{Jackson1975, Mason1979, Sykes1980, Hunt1988, Belcher1998}. Experiments in flumes \citep{Zilker1977, Zilker1979, Frederick1988, Poggi2007}, in wind tunnel \citep{Gong1989, Finnigan1990, Gong1996} and in field conditions \citep{Taylor1987a, Claudin2013, Fernando2019, Lu2021}, have also documented this effect. Interestingly, a similar behaviour exists for the pressure perturbation, but with a slight downwind shift for the pressure minimum \citep{Claudin2021}. The second effect is the flow deflection that occurs when the incident wind direction is not perpendicular to the ridge crest. While predicted to be small (less than $10^{\circ}$) in the linear regime valid for shallow topography \citep{Gadal2019}, significant flow steering has been reported in the field on the downwind side of steep enough obstacles, such as mountain ranges \citep{Kim2000, Lewis2008, Fernando2019} or well developed sand dunes \citep{Walker2009, Hesp2015, Walker2017, Smith2017}.

For practical reasons, wind measurement over sand dunes have been performed over rather small bedforms, typically a few meters high only -- that is tens of meters in length. Giant dunes, with kilometer-scale wavelengths and heights of tens of meters, are more difficult to investigate although they provide, for several reasons, a choice configuration for the study of turbulent flows over a complex topography. First of all, one expects larger wind disturbances for larger obstacles. Second, their large size make them interact with the vertical structure of the atmosphere \citep{andreotti2009}. Also, they usually form large patterns in sand seas and thus behave as rather clean periodic perturbations, in contrast with isolated hills. Finally, because the morphodynamics of aeolian bedforms are strongly dependent on the local wind regime \citep{Livingstone1996}, one can expect to see the consequences of the wind disturbances on smaller dunes, in a similar way as what has been reported for the effect of dunes on impact ripples \citep{Howard1977,Hood2021}.

Arid areas have been much studied at the desert scale from climate reanalyses based on global atmospheric models \citep{Blumberg1996, Livingstone2010, Ashkenazy2012, Jolivet2021, Hu2021}, such as ERA-40, ERA-Interim or ERA-5 \citep{Uppala2005, Dee2011, Hersbach2020}. However, the spatial resolution (tens of kilometers) of these reanalyses implies average quantities that do not resolve the smaller scales, ranging from individual dunes to the border of mountains \citep{Livingstone2010}. Lately, the release of ERA5-Land allows to push back this limitation by providing up to 70 years of hourly wind predictions at a $9$~km spatial resolution \citep{munoz2021}. However, its validity remains to be studied, especially in areas where data assimilation is low (such as deserts).

In this work, we compare local wind speeds and directions measured in four different places inside and north of the giant-dune field of the Namib sand sea to the regional predictions of the ERA5-Land climate reanalysis. When the meteorological stations are surrounded by a relatively flat environment, we show that local measurements and regional predictions agree well with each other. The find that the agreement is also good in the interdune areas of the sand sea, except for the weak winds blowing at night in a direction oblique to the orientation of the giant dunes (section 2). Furthermore, we are able to link the magnitude of these differences to the circadian cycle of the atmospheric boundary layer (section 3). Finally, we draw implications of the wind disturbances on smaller-scale dunes (section 4).

%---------------------------------------------
\begin{figure}[t]
\centering
\includegraphics[scale=1]{Figures/Figure1.pdf}
\caption{Wind data used in this study \textbf{a}: Location of the different sites in Namibia. \textbf{b-e}: Satellite images of these different environements (Google-Earth, Maxar Technologies, CNES/Airbus). In each subplot, the left and right wind roses represent the data from the ERA5-Land climate reanalysis and the local wind stations, respectively. Note that the bars show the direction towards which the wind blows. The black dots show the location of local wind stations.}
\label{Fig1}
\end{figure}
%---------------------------------------------

%---------------------------------------------------------------------------------------
\section{Wind regimes across the Namib Sand Sea}
We have measured the wind regime in four different sites of Namibia, representative of various environments across and nearby the Namib desert (Fig.~\ref{Fig1}). The Adamax station is located near the Adamax salt pan, in a highly vegetated area. The Huab station, on the coast at the outlet of the Huab river, is in an arid place where $60$-m scale barchan dunes develop. These two stations are located in a relatively flat \pc{quantify how flat is flat ?} environment. In contrast, the Deep Sea and South Namib stations are located in the interdune between tens of meters high giant linear dunes with kilometer-scale wavelengths and superimposed patterns. In this section, we describe and compare winds from local measurements and climate reanalysis predictions.

%------------------------
\subsection{Wind and elevation data}
Local winds are measured at meteorological stations located in these four different places (black dots in Fig.~\ref{Fig1}). The wind strength and direction are sampled every 10 minutes by cup anemometers and wind vanes, at heights between $2$~m and $3$~m depending on the station. The available period of measurements ranges from 1 to 5 discontinuous years distributed between 2012 and 2020 (Suppl. Fig.~\ref{Fig1_supp}). We checked that at least one complete seasonal cycle is available at each station. Regional winds are extracted at the same locations and periods from the ERA5-Land dataset, which is a replay at a smaller spatial resolution of ERA5, the latest climate reanalysis from the ECMWWF \citep{Hersbach2020, munoz2021}. It provides hourly predictions of the 10-m wind velocity and direction at a spatial resolution of $0.1^\circ\times0.1^\circ$ ($\simeq 9~\textup{km}$ in Namibia).

For comparison, the local measurements are averaged into 1-hr bins centered on the temporal scale of the ERA5-Land estimates (Suppl. Fig.~\ref{Fig2_supp}). As the wind velocities of both datasets are provided at different heights, we convert them into shear velocities (Suppl. Mat. section 1), characteristic of the turbulent wind profile, which are then used together with the wind direction for further analysis. Wind roses in Fig.~\ref{Fig1}(b--e) show the resulting wind data.

Dune properties are computed using autocorrelation on the 30-m Digital Elevation Models (DEMs) of the shuttle radar topography mission \citep{Farr2007}. For the South Namib and Deep Sea stations, we obtain respectively orientations of $85^\circ$ and $125^\circ$ with respect to the North, wavelengths of $2.6~\textup{km}$ and $2.3~\textup{km}$ and amplitudes of $45~\textup{m}$ and $20~\textup{m}$ (Suppl. Fig.~\ref{Fig4_supp} for more details).

%---------------------------------------------
\begin{figure}
\centering
\includegraphics[scale=1]{Figures/Figure2.pdf}
\caption{Temporal comparison between the wind data coming from the ERA5-Land climate reanalysis (orange lines) and from the local measurements (blue lines). Coloured swathes indicate day (between 1000 UTC and 2200 UTC) and night (before 1000 UTC or after 2200 UTC) \textbf{a-b}: Huab station. \textbf{c-d}: Deep Sea station in winter. \textbf{e-f}: Deep Sea station in summer.}
\label{Fig2}
\end{figure}
%---------------------------------------------

%------------------------
\subsection{Comparison of local and regional winds}
\label{section_data_feedback}
The obtained wind regimes are shown in figure~\ref{Fig1}. In the Namib, the regional wind patterns are essentially controlled by the see breeze, resulting in strong northward components (sometimes slightly deviated by the large scale topography \pc{mountain?}) present in all regional wind roses \citep{lancaster1985}. These daytime winds are dominant during the second-half of the year (September-January). In winter, an additional easterly component can be recorded during the night, induced by the combination of katabatic winds forming in the mountains, and infrequent `berg' winds, which are responsible of the high wind velocities observed \citep{lancaster1984}. The frequency of these easterly components decreases from the inland to the coast. As a result, bidirectional wind regimes within the Namib Sand Sea and at the Adamax salt pan (Fig.~\ref{Fig1}b,d,e) and a unidirectional wind regime on the coast at the outlet of the Huab River (Fig.~\ref{Fig1}c) are observed.

In the case of the Adamax and Huab stations, the time series of wind speed and direction from the regional predictions quantitatively match those corresponding to the local measurements (Fig.~\ref{Fig2}a,b) and Suppl. Fig.~\ref{Fig5_supp}). For the Deep Sea and South Namib stations within the giant dune field, we observe that this agreement is also good, but limited to the September-January time period (Fig.~\ref{Fig2}c,d). As a matter of fact, the measured wind roses exhibit additional wind components aligned with the giant dune orientation, as evidenced on the satellite images (Fig.~\ref{Fig1}c,d).

More precisely, during the February-August period, the local and regional winds in the interdune match during daytime only, i.e when the southerly/southwesterly sea breeze dominates (Figs.~\ref{Fig2}(e,f) and \ref{Fig3}, Suppl. Fig.~\ref{Fig6_supp}). In the late afternoon and during the night, when the northwesterly `berg' and katabatic winds blow, measurements and predictions differ. In this case, the angular wind distribution of the local measurements exhibits two additional modes corresponding to reversing winds aligned with the giant dune orientation (purple frame in Fig.~\ref{Fig3}, Suppl. Figs.~\ref{Fig6_supp} and \ref{Fig7_supp}). This deviation is also associated with a global attenuation of the wind strength (Suppl. Fig.~\ref{Fig8_supp}). Remarkably, all these figures show that this wind reorientation and attenuation processes typically occur at low wind velocities, typically for $u_{*} \lesssim 0.1~\textrm{m}~\textrm{s}^{-1}$. For shear velocities larger than $\simeq 0.25~\textrm{m}~\textrm{s}^{-1}$, the wind reorientation is not significant. Finally, for intermediate shear velocities, moments of reorientation along the dune crest and others with no reorientation can both be observed (Suppl. Fig.~\ref{Fig7_supp}). \pc{CHECK the above lines cf. $\delta u$}

%---------------------------------------------
\begin{figure}
\centering
\includegraphics[scale=1]{Figures/Figure3.pdf}
\caption{Distributions of wind direction at the Deep Sea Station for the ERA5-Land climate reanalysis (orange) and the local measurements (blue). In each subplot, both distributions are plotted from the same time steps, selected with constraints on the wind direction (columns) and/or wind velocity (rows) of the ERA5-Land dataset. The gray vertical dashed lines indicate the dune orientation. The numbers at the top right give the percentage of time steps selected in each sub-range, as well as the percentage of them corresponding to the day (between 1000 UTC and 2200 UTC). The purple frame highlights the regime (small wind velocities, nocturnal summer wind) in which the data from both datasets differ. A similar figure can be obtained for the Deep Sea station (Suppl. Fig.~\ref{Fig6_supp}).}
\label{Fig3}
\end{figure}
%---------------------------------------------

%---------------------------------------------------------------------------------------
\section{Influence of wind speed and circadian cycle on the atmospheric boundary layer}
The wind deflection induced by linear dunes has so far mainly been related to the angle between wind direction and crest orientation, with a maximum deflection for angles between $30^{\circ}$ and $70^{\circ}$ \citep{Walker2009, Hesp2015}. In the analysed data, the most deflected wind at both Deep Sean and South Namib stations is the most perpendicular to the giant dunes (\pc{Fig.~\ref{***}}). The incident wind direction then does not seem to be the only parameter controlling the wind deflection. In contrast, a different behaviour is observed between low and high wind velocities, suggesting a change in hydrodynamical regime. In this section, we discuss the relevant parameters associated with the dynamical mechanisms that govern the interactions between the atmospheric boundary layer flow and topographical obstacles. This analysis allows us to provide a physics-based interpretation the observations.

%---------------------------------------------
\begin{figure}
\centering
\includegraphics[scale=1]{Figures/Figure4.pdf}
\caption{\textbf{a-b}: Vertical profiles of the virtual potential temperature at 2 different time steps (day - 03/11/2015 - 1200 UTC, night - 01/13/2013 - 0900 UTC) at the Deep Sea station. Dots: data from the ERA5 reanalysis. Dashed lines: boundary layer height given by the ERA5 reanalysis (Suppl. Mat. section 2). Plain lines: vertical (boundary layer) and linear (free atmosphere) fits to estimate the stratification properties. \textbf{c-f}: Sketches representing the interaction between the giant dunes and the atmospheric flow for different meteorological conditions. \textbf{g}: Streamlines over a sinusoidal topography $\xi(x,y)$ qualitatively representing the effect of low, medium and strong flow confinement, in relation to the above panels (see Appendix~1 for more details). \pc{add arrow with reference wind}}
\label{Fig4}
\end{figure}
%---------------------------------------------

%------------------------
\subsection{Flow over a modulated bed}
\label{theoretical_framework}
Taking as a reference the turbulent flow over a flat bed, the general framework of this study is the understanding and the description of the flow response to a bed modulation. Without loss of generality, we can consider in this context an idealised bed elevation in the form of parallel sinusoidal ridges, with wavelength $\lambda$ (or wavenumber $k = 2\pi/\lambda$) and amplitude $\xi_0$, and where the reference flow direction makes a given angle with respect to the ridge crest \citep{Andreotti2012}. Part of this response, on which we focus here, is the flow deflection by the ridges. In a simplified way, it can be understood from the Bernoulli principle~\citep{Hesp2015}: as the flow approaches the ridge crest, the compression of the streamlines results in larger flow velocities, and thus lower pressures \citep{Rubin1987}. An incident flow oblique to the ridge is then deflected towards lower pressure zones, i.e towards the crest. Turbulent dissipation tends to increase this effect downstream, resulting in along the crest wind deflection in the lee side~\citep{Gadal2019}.

>>>>

% \title{Local wind regime induced by giant linear dunes}
\title{Local wind regime induced by giant linear dunes: comparison of  ERA5 re-analysis with surface measurements}

@@

\institute{
C. Gadal \at
Institut de M\'ecanique des Fluides de Toulouse, Universit\'e de Toulouse Paul Sabatier, CNRS, Toulouse INP-ENSEEIHT, Toulouse, France. \email{cyril.gadal@imft.fr}
\and
P. Delorme \at
School of Geography and Environmental Science, University of Southampton, Southampton, UK.
\and
C. Narteau \at
Institut de Physique du Globe de Paris, Universit\'e de Paris, CNRS, Paris, France.
\and
G. Wiggs \at
School of Geography and the Environment, University of Oxford, Oxford, UK.
\and
M. Baddock \at
Geography and Environment, Loughborough University, Loughborough, UK.
\and
J.M. Nield \at
School of Geography and Environmental Science, University of Southampton, Southampton, UK.
\and
P. Claudin \at
Physique et M\'ecanique des Milieux H\'et\'erog\`enes, CNRS, ESPCI Paris, PSL Research University, Universit\'e de Paris, Sorbonne Universit\'e, Paris, France.  \email{philippe.claudin@espci.fr}
}

@@

\begin{abstract}
Emergence and growth of sand dunes results from the dynamic interaction between topography, wind flow and sediment transport. While feedbacks between these variables are well studied at the scale of a single dune, the average effect of a periodic dune pattern on atmospheric flows remains poorly constrained due to a lack of data in major sand seas. Here, we compare field measurements of surface wind data to the predictions of the ERA5-Land climate reanalysis at four locations in Namibia, including within the giant-dune field of the Namib sand sea. In the desert plains to the north of the sand sea, observations and predictions agree well. This is also the case in the interdune areas of the sand sea, except for the weak winds blowing at night, which exhibit additional components aligned with the giant dune orientation, which are not predicted by the ERA5-Land reanalysis. We quantify these similarities and differences and provide a physical understanding of the relevant aerodynamical regimes to relate them to the daily cycle of the turbulent atmospheric boundary layer over a dune pattern of given wavelength. We conclude by identifying the conditions under which the ERA5-Land reanalysis data can reliably be used to study dune morphodynamics. We also propose that, in multidirectional wind regimes, deflections of specific winds by giant dunes could explain the occurrence of secondary dune patterns with a different orientation to the primary structures between which they develop.
\keywords{Atmospheric boundary layer \and Sand dunes \and Fluid-structure interactions}
\end{abstract}

@@

The description of turbulent flows over a complex topography is a rich and active subject, that is relevant for a large variety of different environmental systems \citep{Finnigan2020}. For example, the flow over hills is of primary interest for wind power, meteorological and air pollution phenomena \citep{Taylor1987}. The properties of these flows are also key to the understanding of the formation of ocean surface wind-driven waves \citep{Sullivan2010}, dissolution bedforms \citep{Claudin2017}, or sedimentary ripples and dunes \citep{Charru2013, Courrech2015}. Importantly, the troposphere presents a vertical structure, with a lower convective boundary layer, of typical kilometer-scale thickness, capped by a stably stratified region \citep{Stull1988}. The largest topographic obstacles, such as mountains can therefore interact with this upper region, and lead to internal wave generation or significant wind disturbances, such as lee-side downslope winds \citep{Durran1990}.

Focusing on the wind close to the surface, two related topographic feedbacks on the windflow over dunes can be commented on separately. First is the effect on wind speed with documented flow acceleration on upwind slopes \citep{Weaver2011} and deceleration on downwind slopes \citep{Baddock2007}, with a speed-up factor essentially proportional to the obstacle aspect ratio \citep{Jackson1975}. Importantly, the velocity maximum is typically shifted upwind of the obstacle crest. This behaviour has been theoretically predicted by means of asymptotic analysis of a neutrally stratified boundary-layer flow over an obstacle of vanishing aspect ratio \citep{Jackson1975, Mason1979, Sykes1980, Hunt1988, Belcher1998}. Experiments in flumes \citep{Zilker1977, Zilker1979, Frederick1988, Poggi2007}, in wind tunnels \citep{Gong1989, Finnigan1990, Gong1996} and in field conditions \citep{Taylor1987a, Claudin2013, Fernando2019, Lu2021, Bristow2022}, have also documented this effect. Interestingly, a similar behaviour exists for the pressure perturbation, but with a slight downwind shift for the pressure minimum \citep{Claudin2021}. The second effect, much less studied, is the flow deflection that occurs when the incident wind direction is not perpendicular to the ridge crest. While predicted to be small (less than $10^{\circ}$) in the linear regime valid for shallow topography \citep{Gadal2019}, significant flow steering has been reported in the field on the downwind side of steep enough obstacles, such as mountain ranges \citep{Kim2000, Lewis2008, Fernando2019}, well-developed sand dunes \citep{Walker2009, Hesp2015, Walker2017, Smith2017}, and valley topographies \citep{Wiggs2002}.

For practical reasons, wind measurement over sand dunes has been performed over rather small bedforms, typically a few meters high only (corresponding to tens of meters long) \citep{Wiggs1996}. Giant dunes, with kilometer-scale wavelengths and heights of tens of meters, are more difficult to investigate although for several reasons they provide a choice configuration for the study of turbulent flows over a complex topography. First, one expects larger wind disturbances for larger obstacles. Secondly, their large size makes them interact with the vertical structure of the atmosphere \citep{Andreotti2009}. Third, they usually form large patterns in sand seas and thus behave as rather clean periodic perturbations, in contrast with isolated dunes. Finally, because the morphodynamics of aeolian bedforms are strongly dependent on the local wind regime \citep{Livingstone2019}, one can expect to see the consequences of windflow disturbance by large dunes on neighbouring small dunes, in a similar manner to that reported for the effect of dunes on impact ripples \citep{Howard1977,Hood2021}.

Atmospheric flows have been much studied at the desert-scale with climate reanalyses based on global atmospheric models \citep{Blumberg1996, Livingstone2010, Ashkenazy2012, Jolivet2021, Hu2021}, such as ERA-40, ERA-Interim or ERA-5 \citep{Uppala2005, Dee2011, Hersbach2020}. However, the spatial resolution (tens of kilometers) of these reanalyses implies average quantities that do not resolve the smaller scales, ranging from individual dunes to the border of mountains \citep{Livingstone2010}. Recently, the release of ERA5-Land has resolved this limitation by providing up to 70 years of hourly wind predictions at a $9$~km spatial resolution \citep{munoz2021}. However, its validity remains to be studied, especially in remote desert areas where data assimilation is very low.

In this work, we compare local wind speeds and directions measured by meteorological stations at four different locations inside and north of the giant-dune field of the Namib sand sea to the regional predictions of the ERA5-Land climate reanalysis. Where the meteorological stations are surrounded by a relatively flat environment, we show that local measurements and regional predictions agree well. The agreement is also good in the interdune areas of the sand sea, except for some weak winds blowing at night, which exhibit an additional component aligned with the giant dune orientation, which are not predicted by the ERA5Land reanalysis (section 2). Further, we are able to link the magnitude of these differences to the circadian cycle of the atmospheric boundary layer (section 3). Finally, we draw implications for the wind disturbances on smaller-scale dunes (section 4), suggesting a possible origin for crossing dunes.

%---------------------------------------------
\begin{figure}[t]
\centering
\includegraphics[scale=1]{Figures/Figure1.pdf}
\caption{Wind data used in this study \textbf{a}: Location of the different sites in Namibia. \textbf{b--e}: Satellite images of these different environements (Google-Earth, Maxar Technologies, CNES/Airbus). In each subplot, the left and right wind roses represent the data from the ERA5-Land climate reanalysis and the local wind stations, respectively. Note that the bars show the direction towards which the wind blows. The black dots show the location of local wind stations.}
\label{Fig1}
\end{figure}
%---------------------------------------------

%---------------------------------------------------------------------------------------
\section{Wind regimes across the Namib Sand Sea}
We measured the wind regime at four different locations in Namibia, representative of various arid environments across the Namib desert (Fig.~\ref{Fig1}, Online Resource Fig.~\ref{Fig1_supp}). The Adamax station was located at the Adamax waterhole to the west of Etosha Pan in northern Namibia, in a sparsely vegetated area. The Huab station was near the coast on a hyper-arid flat gravel plain lying above the ephemeral Huab river. Here, 5--6~m high barchan dunes develop from the sediment blowing out of the river valley. These two stations were located in a relatively flat environments. In contrast, the North Sand Sea and South Sand Sea stations were located in the interdunes between 50--100~m high linear dunes with kilometer-scale wavelengths and superimposed patterns. In this section, we describe and compare winds from local measurements and climate reanalysis predictions.

%------------------------
\subsection{Wind and elevation data}
At each meteorological station (Fig.~\ref{Fig1}), wind speed and direction were sampled every 10 minutes using cup anemometers and wind vanes at heights between 2~m and 3~m depending on the station. The available period of measurements at each station ranged from 1 to 5 discontinuous years distributed between 2012 and 2020 (Online Resource Fig.~\ref{Fig2_supp}). We checked that at least one complete seasonal cycle was available for each station. Regional winds were extracted at the same locations and periods from the ERA5-Land dataset, which is a replay at a smaller spatial resolution of ERA5, the latest climate reanalysis from the ECMWWF \citep{Hersbach2020, munoz2021}. This dataset provided hourly predictions of the 10-m wind velocity and direction at a spatial resolution of $0.1^\circ\times0.1^\circ$ ($\simeq 9~\textup{km}$ in Namibia).

To enable direct comparison, the local wind measurements were averaged into 1-hr bins centered on the temporal scale of the ERA5-Land estimates (Online Resource Fig.~\ref{Fig3_supp}). As the wind velocities of both datasets were provided at different heights, we converted them into shear velocities $u_{*}$ (Online Resource section 1), characteristic of the turbulent wind profile. Wind roses in Fig.~\ref{Fig1}(b--e) show the resulting wind data.

Dune properties were computed using autocorrelation on the 30-m Digital Elevation Models (DEMs) of the shuttle radar topography mission \citep{Farr2007}. For the South Sand Sea and North Sand Sea stations, we obtain, respectively, orientations of $85^\circ$ and $125^\circ$ with respect to the North, wavelengths of $2.6~\textup{km}$ and $2.3~\textup{km}$ and amplitudes (or half-heights) of $45~\textup{m}$ and $20~\textup{m}$ (Online Resource Fig.~\ref{Fig5_supp} for more details).

%---------------------------------------------
\begin{figure}
\centering
\includegraphics[scale=1]{Figures/Figure2.pdf}
\caption{Temporal comparison between the wind data coming from the ERA5-Land climate reanalysis (orange lines) and from the local measurements (blue lines). Coloured swathes indicate day (between 1000 UTC and 2200 UTC) and night (before 1000 UTC or after 2200 UTC) \textbf{a--b}: Adamax station in summer. \textbf{b--c}: Adamax station in winter. \textbf{d--e}: North Sand Sea station in summer. \textbf{f--g}: North Sand Sea station in winter. Time series of the two other stations are shown in Online Resource Fig.~\ref{Fig6_supp}.}
\label{Fig2}
\end{figure}
%---------------------------------------------

%------------------------
\subsection{Comparison of local and regional winds}
\label{section_data_feedback}
The obtained measured and predicted wind regimes are shown in figure~\ref{Fig1}. In the Namib, the regional wind patterns are essentially controlled by the see breeze, resulting in strong northward components (sometimes slightly deviated by the large scale topography) present in all regional wind roses \citep{lancaster1985}. These daytime winds are dominant during the period October-March (Fig.~\ref{Fig2}f and Online Resource Fig.~\ref{Fig6_supp}f). During April-September, an additional easterly component can be recorded during the night, induced by the combination of katabatic winds forming in the mountains, and infrequent `berg' winds, which are responsible for the high wind velocities observed \citep{lancaster1984}. The frequency of these easterly components decreases from inland to the coast. As a result, bidirectional wind regimes within the Namib Sand Sea and at the Adamax salt pan (Fig.~\ref{Fig1}b,d,e) and a unidirectional wind regime on the coast at the outlet of the Huab River (Fig.~\ref{Fig1}c) are observed.

In the case of the Adamax and Huab stations, the time series of wind speed and direction from the regional predictions quantitatively match those corresponding to the local measurements (Fig.~\ref{Fig2}a--d) and Online Resource Fig.~\ref{Fig6_supp}a--d,~\ref{Fig7_supp}). For the North Sand Sea and South Sand Sea stations within the giant dune field, we observe that this agreement is also good, but limited to the October-March time period (Fig.~\ref{Fig2}e--h and and Online Resource Fig.~\ref{Fig6_supp}e--h). As a matter of fact, the measured wind roses exhibit additional wind components aligned with the giant dune orientation, as evidenced on the satellite images (Fig.~\ref{Fig1}c,d).

More precisely, during the April-September period, the local and regional winds in the interdune match during daytime only, i.e when the southerly/southwesterly sea breeze dominates (Figs.~\ref{Fig2}e,f and \ref{Fig3}, Online Resource Fig.~\ref{Fig8_supp}). In the late afternoon and during the night, when the northwesterly `berg' and katabatic winds blow, measurements and predictions differ. In this case, the angular wind distribution of the local measurements exhibits two additional modes corresponding to reversing winds aligned with the giant dune orientation (purple frame in Fig.~\ref{Fig3}, Online Resource Figs.~\ref{Fig8_supp} and \ref{Fig9_supp}). This deviation is also associated with a general attenuation of the wind strength (Online Resource Fig.~\ref{Fig10_supp}). Remarkably, all these figures show that these wind reorientation and attenuation processes occur only at low wind velocities, typically for $u_{*} \lesssim 0.1~\textrm{m}~\textrm{s}^{-1}$. For shear velocities larger than $\simeq 0.25~\textrm{m}~\textrm{s}^{-1}$, the wind reorientation is not apparent. Finally, for intermediate shear velocities, both situations of wind flow reoriented along the dune crest and not reoriented can be successively observed (Online Resource Fig.~\ref{Fig9_supp}).

%---------------------------------------------
\begin{figure}
\centering
\includegraphics[scale=1]{Figures/Figure3.pdf}
\caption{Distributions of wind direction at the North Sand Sea Station for the ERA5-Land climate reanalysis (orange) and the local measurements (blue). In each subplot, both distributions are plotted from the same time steps, selected with constraints on the wind direction (columns) and/or wind velocity (rows) of the ERA5-Land dataset. The grey vertical dashed lines indicate the dune orientation. The numbers at the top right give the percentage of time steps selected in each sub-range, as well as the percentage corresponding to the daytime (between 1000 UTC and 2200 UTC). The purple frame highlights the regime (low wind velocities, nocturnal easterly wind) in which the data from both datasets differ. A similar figure can be obtained for the North Sand Sea station (Online Resource Fig.~\ref{Fig8_supp}).}
\label{Fig3}
\end{figure}
%---------------------------------------------

%---------------------------------------------------------------------------------------
\section{Influence of wind speed and circadian cycle on the atmospheric boundary layer}
The wind deflection induced by linear dunes has previously been related to the incident angle between wind direction and crest orientation, with a maximum deflection evident for incident angles between $30^{\circ}$ and $70^{\circ}$ \citep{Walker2009, Hesp2015}. In the data analysed here, the most deflected wind at both the North and South Sand Sea stations is seen to be where the incident angle is perpendicular to the giant dunes (Figs.~\ref{Fig1} and \ref{Fig3}, Online Resource Fig.~\ref{Fig8_supp}). It therefore appears that the incident wind angle is not the dominant control on maximum wind deflection. Further, and as shown in Figure~\ref{Fig3}, winds of high and low velocities show contrasting behaviour in characteristics of deflection. This suggests a change in hydrodynamical regime between the winds. In this section, we discuss the relevant parameters associated with the dynamical mechanisms that govern the interactions between the atmospheric boundary layer flow and giant dune topographies. This analysis allows us to provide a physics-based interpretation of our measured wind data.

%---------------------------------------------
\begin{figure}
\centering
\includegraphics[scale=1]{Figures/Figure4.pdf}
\caption{\textbf{a--b}: Vertical profiles of the virtual potential temperature at 2 different time steps (day - 03/11/2015 - 1200 UTC, night - 01/13/2013 - 0900 UTC) at the North Sand Sea station. Dots: data from the ERA5 reanalysis. Dashed lines: boundary layer height given by the ERA5 reanalysis (Online Resource section 2). Plain lines: vertical (boundary layer) and linear (free atmosphere) fits to estimate the stratification properties. \textbf{c--f}: Sketches representing the interaction between the giant dunes and the atmospheric flow for different meteorological conditions. \textbf{g}: Streamlines over a sinusoidal topography $\xi(x,y)$ qualitatively representing the effect of low, medium and strong flow confinement, in relation to the above panels (see Appendix~1 for more details). The blue arrow indicates the undisturbed wind direction.}
\label{Fig4}
\end{figure}
%---------------------------------------------

%------------------------
\subsection{Flow over a modulated bed}
\label{theoretical_framework}
Taking as a reference the turbulent flow over a flat bed, the general framework of our study is understanding and describing the flow response to a bed modulation (e.g. a giant dune). Without loss of generality, we can consider in this context an idealised bed elevation in the form of parallel sinusoidal ridges, with wavelength $\lambda$ (or wavenumber $k = 2\pi/\lambda$) and amplitude $\xi_0$, and where the reference flow direction makes a given incident angle with respect to the ridge crest \citep{Andreotti2012}. Part of this response, on which we focus here, is the flow deflection by the ridges. In a simplified way, it can be understood from the Bernoulli principle~\citep{Hesp2015}: as the flow approaches the ridge crest, the compression of the streamlines results in larger flow velocities, and thus lower pressures \citep{Rubin1987}. An incident flow oblique to the ridge is then deflected towards lower pressure zones, i.e towards the crest. Turbulent dissipation tends to increase this effect downstream, resulting in wind deflection along the crest in the lee side~\citep{Gadal2019}.
